---
title: "`BulkSignalR` : </br> Inference of ligand-receptor interactions from bulk data or spatial transcriptomics"
author:
  - name: Jean-Philippe Villemin
    affiliation:
    - Institut de Recherche en Cancérologie de Montpellier, Inserm, Montpellier, France 
    email: jean-philippe.villemin@inserm.fr
  - name: Jacques Colinge
    affiliation:
    - Institut de Recherche en Cancérologie de Montpellier, Inserm, Montpellier, France 
    email: jacques.colinge@inserm.fr
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output: 
  rmarkdown::html_vignette:
      self_contained: true
      toc: true
      toc_depth: 4
      highlight: pygments
      fig_height: 3
      fig_width: 3
      fig_caption: no
      code_folding: show
vignette: >
  %\VignetteIndexEntry{BulkSignalR-Differential}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE, 
  comment = "##"
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

BulkSignalR package version: `r packageVersion("BulkSignalR")`

## Differential mode 

`BulkSignalR` now allows the user to request different pairwise
comparisons from a large cohort of samples.
  
In this mode, ligand-receptor interactions are inferred based on gene or protein
regulation-associated P-values when comparing two clusters of samples. 

The user must first perform a differential expression analysis 
(using the tool of his choice as DESeq2, EdgeR...) between a 
pair of sample clusters.


In the next chunk of code, we describe an application to 
Salivary Duct Carcinoma (SDC) samples where we compare two clusters 
of patients.

```{r diffmode,eval=FALSE}
library(BulkSignalR)

data(sdc,package='BulkSignalR')
normal <- grep("^N", names(sdc))
bsrdm <- prepareDataset(sdc[,-normal])

# define the comparison

bsrdm.comp <- as.BSRDataModelComp(bsrdm)
colA <- as.integer(1:5)
colB <- as.integer(8:15)

# As an example here, we generate random values 
# but user should provide his own logFC and
# associated pvalues from DGE ouputs. 

n <- nrow(ncounts(bsrdm.comp))
stats <- data.frame(pval=runif(n), logFC=rnorm(n, 0, 2))
rownames(stats) <- rownames(ncounts(bsrdm.comp))

# we first define the cluster comparison and add it
# to the BSRDataModelComp object.

bsrcc <- defineClusterComp(bsrdm.comp, colA, colB, stats)
bsrdm.comp <- addClusterComp(bsrdm.comp, bsrcc, "random.example")

# finally we infer ligand-receptor interactions from the comparison

bsrinf.comp <- initialInference(bsrdm.comp, max.pval=1,"random.example")


```  

##  Technical notes

Three previously described S4 objects (BSR-DataModel, BSR-Inference, BSR-Signature) 
have been extented :  

\
  

* **BSR-DataModelComp**, denoted `bsrdm.comp` is an extension from 
**BSR-DataModel**, previously denoted `bsrdm`  
* **BSR-InferenceComp**, denoted `bsrinf.comp` is an extension from 
**BSR-Inference**, previously denoted `bsrinf` 
* **BSR-SignatureComp**, denoted `bsrsig.comp` is an extension from 
**BSR-Signature**, previously denoted `bsrsig` 

A new S4 object **BSR-ClusterComp** representing the comparison of two clusters
of samples to infer LR interactions based on the resulting P-values and
log-fold-changes (logFC) has been added.

This new desgin let us handle sample cluster comparisons.


`as.BSRDataModelComp` converts of **BSR-DataModel** into **BSR-DataModelComp** 
In case ligand-receptor inferences should be obtained
based on gene/protein regulation P-values comparing two clusters of
samples, it is necessary to first promote the **BSR-DataModel** object that
contains the count matrix into a **BSR-DataModelComp** object able to contain
a list of such cluster pairs comparisons. This function performs this
promotion, adding an empty list of comparisons.  

`defineClusterComp` is used to define the comparison between two clusters of samples
by using defined column indexes of the expression matrix that belong to each cluster,
and storing the result of the cluster difference statistical analysis
obtained by an external tool such as edgeR, DESeq2, etc.

`addClusterComp` add a comparison between two clusters of samples 
to a **BSR-DataModelComp** object.

Several comparison can be defined and added sequentially.  

\
 
Thank you for reading this guide and for using `BulkSignalR`.  

## Session Information

```{r session-info}
sessionInfo()
```
